### 2023-02-21 tex/article.tex
Add gls for Chl

% \begin{keypoints}
% \item Over SST fronts in the Gulf Stream region Chlorophyll\nobreakdash-\emph{a} is enhanced by up to a factor two and the bloom onsets earlier by up to two weeks.
% \item The enhancement intensity increases significantly with the strength of the front.
% \item The net contribution of fronts to Chlorophyll\nobreakdash-\emph{a} enhancement over the Gulf Stream region is less than 5\%.
% \end{keypoints}

\begin{abstract}
Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing vertical nutrient supply.
Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling, and reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
However observation of these biophysical interactions through satellite imagery have been scarce, despite the opportunity to quantify them at synoptic scales.
Here we used twenty years of SST and Chlorophyll\nobreakdash-\emph{a} satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on phytoplankton in contrasting regimes, from oligotrophy to bloom, and throughout the year.
We computed an heterogeneity index HI from SST, and used it to sort fronts into weak and strong fronts based on HI thresholds.
We observed that the localization of strong fronts corresponded to western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
We compared Chlorophyll\nobreakdash-\emph{a} distributions over strong fronts, weak fronts and outside of fronts.
We assessed three metrics, the local enhancement of Chlorophyll\nobreakdash-\emph{a} over fronts, the global enhancement of Chlorophyll\nobreakdash-\emph{a} due to fronts at the scale of the region, and the lag in spring bloom onset due to fronts.
We found that weak fronts lead to a local enhancement of Chlorophyll\nobreakdash-\emph{a} weaker than strong fronts, but because they are also more frequent they contribute equally to the regional Chlorophyll\nobreakdash-\emph{a} budget.
We also find the the local enhancement of Chlorophyll\nobreakdash-\emph{a} was two to three times larger for the spring bloom than in the oligotrophic subtropical gyre.
We also provide observational evidence that blooms start earlier over fronts, by one to two weeks.
Nevertheless our results suggest that the spectacular impact of fronts at the local scale may be misleading, considering their impact on a regional scale budget remains limited.
\end{abstract}

%\copyrightstatement{TEXT} %% This section is optional and can be used for copyright transfers.

@@

Despite numerous local observations and a strong theoretical basis for these processes \citep[e.g. recent studies by][]{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}, their integrated contribution at the scale of regional biomes is still largely unknown.
Indeed, as most fronts form, move and dissipate continuously on time scales of days to weeks, they are particularly difficult to sample, and this limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of Chlorophyll\nobreakdash-\emph{a} (hereafter Chl\nobreakdash-\emph{a}) although limited to the surface and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of fine scales to regional satellite Chl\nobreakdash-\emph{a} budgets was based on a geostatistical analysis derived from data at 9km resolution \citep{doney_2003}, extended later in \citet{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of submesoscales but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of submesoscales has been assessed with three different methods.
\citet{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscales and submesoscales to high Chl\nobreakdash-\emph{a} anomalies were comparable in magnitude.
\citet{keerthi_2022} proposed an approach based on deconvolution of local Chl\nobreakdash-\emph{a} time series into different timescales; they observed that sub-seasonal time scales contributed roughly 30\% of the total satellite Chl\nobreakdash-\emph{a} variance and were associated with small (\textless\ 100km) spacial scales.
Finally, the most quantitative approach was proposed by \citet{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (SST) fronts by computing an index that measures the local heterogeneity of the SST field from satellite SST data.
This allowed them to compare satellite Chl\nobreakdash-\emph{a} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in Chl\nobreakdash-\emph{a} over the fronts was negligible in summer but reached almost 40\% in winter.

Here we built on this last approach, and quantify the excess Chl\nobreakdash-\emph{a} due to the presence of fronts at the scale of biomes.
This more global quantification will depend on several factors, since the local contribution of fronts to Chl\nobreakdash-\emph{a} depends on many factors.
First it depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, and of the seasonality of this supply.
It should be noted that the overall efficiency of fronts has been questioned \citep{levy_2018}; firstly because submesoscale fronts are more numerous in winter when convective nutrients input is also greatest; and secondly because the submesoscale vertical currents are often trapped within the mixed-layer, and may not reach the nutricline, which is often well below the base of the boundary mixed-layer.
Second, the overall contribution of fronts will differ between biomes, with submesoscale vertical advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the contribution of fronts will depend on their spatio-temporal footprint, which also varies seasonally \citep{callies_2015} and regionally \citep{mauzole_2022}.

Thus, more precisely, our intention is to explore and quantify how the contribution of fronts to biome-scale Chl\nobreakdash-\emph{a} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.
We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area \citep{bock_2022} with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has different types of fronts.
On the one hand, there are two strong persistent fronts, the Gulf Stream and the shelf-break front, which are held in place by topography and atmospheric circulation, and which are both associated with strong and deep-reaching vertical circulations \citep{flagg_2006, liao_2022}.
But there are also more ephemeral and weaker fronts, created by mesoscale strain that are continuously forming at more random locations \citep{drushka_2019, sanchez-rios_2020}.
We use satellite data of Chl\nobreakdash-\emph{a} and SST, and extend on the approach of \citet{liu_2016}, to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on Chl\nobreakdash-\emph{a} on the basis of three indicators, the excess (or deficit) Chl\nobreakdash-\emph{a} over fronts at the local scale of the front, the surplus Chl\nobreakdash-\emph{a} attributable to fronts at the scale of regional biomes, and the change in the timing of the Chl\nobreakdash-\emph{a} spring bloom over fronts.

@@

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface Chl\nobreakdash-\emph{a}, from which we derive anomalies over fronts.
Our region of interest is the North Atlantic from 15°N to 55°N, and from 40°W to the North American shelf break (Fig.~\ref{fig:zone-separation}).
This region covers three biomes, a more oligotrophic one in the south, a more productive one in the north where a spring bloom occurs, and an intermediate biome between the two, described below.
We will refer to them in the following as permanent subtropical biome (PSB), seasonal subtropical biome (SSB) and subpolar biome (PB), moving from south to north.
They are separated from south to north by a fixed boundary at 32°N and by a variable boundary near 40°N dynamically set at each time step to the instantaneous position of the Gulf Stream.

@@

For Chl\nobreakdash-\emph{a}, we used the L3 product distributed by ACRI-ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (SeaWIFS, MODIS Aqua \& Terra, MERIS, VIIRS-SNPP \& JPSS1, OLCI-S3A \& S3B) merged and reprocessed, available daily at 4km resolution \citep{chl}.

@@

The three biomes in our domain correspond to well known production regimes which have been described previously, although sometimes with different names.
The region is characterized by the presence of a large-scale north-south gradient in Chl\nobreakdash-\emph{a}.
All pixels where water depth is less than 1500m (red isobath in Fig.~\ref{fig:zone-separation}) are masked to exclude the continental shelf.

The permanently oligotrophic regime to the south of our study area, characterized by warm waters and low Chl\nobreakdash-\emph{a} (Fig.~\ref{fig:zone-separation}), is known as the subtropical gyre permanently stratified biome \citep{sarmiento_2004} or permanent deep Chl\nobreakdash-\emph{a} maximum biome \citep{bock_2022}.
Their is no clear physical boundary to the northern limit of this permanent subtropical biome, so we have chosen the latitudinal limit of 32°N to delineate it, which roughly corresponds to the \(0.1 mg m^{-3}\) Chl\nobreakdash-\emph{a} isocontour in annual mean Chl\nobreakdash-\emph{a}.

North of 32°N, the seasonal subtropical biome is also mainly oligotrophic, with intermediate levels of Chl\nobreakdash-\emph{a} and temperature, and characterized by slightly increased productivity in winter.
This second biome is known as the subtropical gyre seasonally stratified biome \citep{sarmiento_2004}.
It is bounded to the north by the meanders of the Gulf Stream jet.
The Gulf Stream jet conveys warm, salty waters poleward along the Florida coast up to Cape Hatteras (35°N), where the jet separates from the continental shelf and meanders essentially zonally.
The north wall of the jet, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of this biome.

@@

The position of the north wall of the Gulf stream, that delimits the subpolar and seasonal subtropical biomes (black meandering contour in Fig.~\ref{fig:zone-separation}a-b), is determined at each daily time step by thresholding the daily SST map.
The daily threshold values (black vertical line in Fig.~\ref{fig:zone-separation}c) are determined from the daily SST distributions above 32°N (blue line, continued by the yellow line in Fig.~\ref{fig:zone-separation}c).
Indeed, the Gulf Stream is easily identifiable in this distribution as it is manifested by a temperature peak (yellow line in Fig.~\ref{fig:zone-separation}c).
To detect the start of this peak, which marks our boundary, we fit the peak with a Gaussian, and define the threshold temperature as the mean minus twice the standard deviation.
Then, the threshold temperature time series is median filtered over an 8-day window to eliminate spurious detection anomalies.
This separation method ensures quasi-unimodal Chl\nobreakdash-\emph{a} distributions within each biome (Fig.~\ref{fig:zone-separation}d).

@@

\begin{figure*}
  \includegraphics[width=12cm]{zone_separation.pdf}
  \caption{
    Delimitation of the three biomes in the Gulf stream extension region: the Permanent Subtropical Biome (PSB, south of the dashed line at 32°N), the Seasonal Subtropical Biome (SSB, between 32°N and the meandering Gulf stream northern wall on that day marked with the black contour), and the Subpolar Biome (PB, north of the Gulf stream northern wall).
    (a) SST and (b) Chl\nobreakdash-\emph{a} snapshots on the 22 April 2007 (with data masked by clouds in white), (c) SST and (d) Chl\nobreakdash-\emph{a}  distribution within each biome for the same day (PB:\@blue, SSB:\@yellow, PSB:\@red).
    The black line in (c) shows the SST threshold value detected to delimit the Gulf Stream northern wall (see methods section).
    The x-axis of the distributions correspond to the x-axis scale of the corresponding color bars.
    The red line follows the 1500m isobath.
    Data on the continental shelf (< 1500m) is not considered here and have been masked.
  }%
  \label{fig:zone-separation}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{front_example.pdf}
  \caption{
    The SST, Chl\nobreakdash-\emph{a}, and heterogeneity index (HI) of a front on the 7 July 2007.
    The plain and dashed contours correspond to HI values of 5 and 10.
    This front is categorized as weak.
    Chl\nobreakdash-\emph{a} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

\subsection{Quantification of the impact of fronts on Chl\nobreakdash-\emph{a}}

We used three metrics to quantify the impact of fronts on Chl\nobreakdash-\emph{a}, the excess Chl\nobreakdash-\emph{a}  E for the local scale (scale of fronts), the surplus Chl\nobreakdash-\emph{a}  S for the global scale (scale of biomes) and the lag in onset day L for the bloom timing.

In order to quantify how Chl\nobreakdash-\emph{a} is affected by fronts at the local scale, the metrics E is based on the comparison between the distributions of Chl\nobreakdash-\emph{a} over fronts and the distributions of Chl\nobreakdash-\emph{a} outside of fronts.
Distributions were computed over 8-day windows to limit the influence of particularly cloudy days.
They were computed within each biome, and over latitudinal bands of width 5°, to minimize the influence of the large-scale north south gradient in Chl\nobreakdash-\emph{a}.
Distributions were compared in terms of their median value, but using mean values yielded similar results (not shown).
We defined and computed the local excess Chl\nobreakdash-\emph{a}, E (expressed in per cent), as the median value over fronts minus the median value in the background, divided by the median value in the background, for each distribution.
We repeated this for weak and strong fronts.

To quantify the large-scale impact of fronts on Chl\nobreakdash-\emph{a} at the scale of biomes, we computed the biome surplus of Chl\nobreakdash-\emph{a} which we defined and computed as the relative difference (expressed in per cent) between the mean Chl\nobreakdash-\emph{a} over the entire biome (MT), and the mean Chl\nobreakdash-\emph{a} over the background (MB), \(S = (\mathrm{MT}-\mathrm{MB}) / \mathrm{MB}\).
Thus the surplus S measures the extra quantity of Chl\nobreakdash-\emph{a} at the scale of the biome (\(\mathrm{MT} - \mathrm{MB}\)), relative to what would be the situation in the absence of fronts (MB), and thus accounts for the local increase over fronts (E), but also for the proportion of fronts in a given biome.
To better understand the meaning of the surplus, let us consider the simplified case where Chl\nobreakdash-\emph{a} is homogeneously doubled over fronts compared to the background value, i.e.\ when E = 100\%; in that case, the surplus Chl\nobreakdash-\emph{a} is 50\% if there are 50\% of fronts, and is 1\% if there are 1\% of fronts.
Note that the computation of the local excess E is based on median values because it relies on the comparison of distributions, while for the computation of the biome surplus S, we used mean values in order to be conservative.

Finally, the time series of the Chl\nobreakdash-\emph{a} median in the subpolar biome is characterized by a spring bloom, of which we measured the timing (onset date) both in the fronts and in the background.
Because the spring bloom onsets propagates from south to north, these timings were inferred over latitudinal band of width 5°.
To extract the onset date, we filtered the Chl\nobreakdash-\emph{a} median time series with a low-pass Butterworth filter of order 2 and cutting frequency \(1/20 \textrm{days}^{-1}\).
The filtered time-series displayed strong variations in their phenology from year to year, but a bloom was always discernible.
We considered data from February to July, which allowed us to isolate the spring bloom and exclude the autumn bloom.
First, we detected the maximum value of Chl\nobreakdash-\emph{a} in this time window, then defined the bloom onset as the time of maximum Chl\nobreakdash-\emph{a} derivative prior to the time of maximum Chl\nobreakdash-\emph{a}.
To estimate the uncertainty in this evaluation, we computed the standard deviation of all days for which the Chl\nobreakdash-\emph{a} derivative was above 90\% of its maximum value.
Finally, we averaged the yearly values of lag L in onset days over the 20 years range by computing a weighted average and standard deviation of the difference between values in fronts and in background for each latitudinal band, with the weights equal to the inverse of the standard deviations for each year.

@@

\begin{figure*}
  \includegraphics[width=12cm]{chl_vs_hi.pdf}
  \caption{
    Normalized distribution of the Heterogeneity Index (HI, red dashed line) within each biome, and distribution of Chl\nobreakdash-\emph{a} as a function of HI (representing front strength), over the full period 2000--2020.
    Shown are the median value of the Chl\nobreakdash-\emph{a} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that 0.5\% of pixels have outstanding large HI values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

@@

\subsection{Local impact of fronts on Chl\nobreakdash-\emph{a}}

The local impact of fronts on Chl\nobreakdash-\emph{a} is illustrated by the example shown in Fig.~\ref{fig:zoom}, where the highest values of Chl\nobreakdash-\emph{a} are found within the HI contour delimiting fronts.
It should be noted, however, that in this example as in other similar examples, the localization of the highest Chl\nobreakdash-\emph{a} do not perfectly coincide with elevated values of HI;\ they are places where the HI is large and Chl\nobreakdash-\emph{a} is small, and there are also places with elevated patches of Chl\nobreakdash-\emph{a} outside of HI contours.

Nevertheless, it appears that on average, Chl\nobreakdash-\emph{a} is affected by the presence of fronts, and depends on their strength.
In order to measure that effect, we computed for each biome the Chl\nobreakdash-\emph{a} distribution sorted by bins of HI (of width 0.1) for the whole time series (Fig.~\ref{fig:chl-vs-hi}).
For low values of the heterogeneity index, these distributions are representative of background conditions and reflect the expected differences between biomes: the median Chl\nobreakdash-\emph{a} is lowest (\(0.05 mg m^{-3}\)) in the permanent subtropical biome, intermediate in the seasonal subtropical biome (\(0.1 mg m^{-3}\)), and highest in the subpolar biome (\(0.2 mg m^{-3}\)).
The Chl\nobreakdash-\emph{a} variability along the year and within the biome (i.e.\ the width of the distribution, highlighted by grey shading in Fig.~\ref{fig:chl-vs-hi}) is larger moving northwards, because of the higher seasonal variability.

Importantly, in all 3 biomes, Chl\nobreakdash-\emph{a} values increase with HI, and are also more dispersed as HI increases (Fig.~\ref{fig:chl-vs-hi}).
This suggests that the distinction between background, weak and strong fronts is somehow artificial and that the observed changes in Chl\nobreakdash-\emph{a} are overall rather continuously dependent on the value of HI.\@
Nevertheless, our partitioning into background, weak and strong fronts remains meaningful because it enables us to isolate the most permanent fronts from the ephemeral ones, and we will therefore retain it in the following.

>>>>

% \begin{keypoints}
% \item Over SST fronts in the Gulf Stream region al{chl} is enhanced by up to a factor two and the bloom onsets earlier by up to two weeks.
% \item The enhancement intensity increases significantly with the strength of the front.
% \item The net contribution of fronts to al{chl} enhancement over the Gulf Stream region is less than 5\%.
% \end{keypoints}

\begin{abstract}
Fronts affect phytoplankton growth and phenology by locally reducing stratification and increasing vertical nutrient supply.
Biomass peaks at fronts have been observed in-situ and linked to local nutrient upwelling, and reduced stratification over fronts has been shown to induce earlier blooms in numerical models.
However observation of these biophysical interactions through satellite imagery have been scarce, despite the opportunity to quantify them at synoptic scales.
Here we used twenty years of SST and al{chl} satellite data in a large region surrounding the Gulf Stream to quantify the impact of fronts on phytoplankton in contrasting regimes, from oligotrophy to bloom, and throughout the year.
We computed an heterogeneity index HI from SST, and used it to sort fronts into weak and strong fronts based on HI thresholds.
We observed that the localization of strong fronts corresponded to western boundary current fronts, and weak fronts to more ephemeral submesoscale fronts.
We compared al{chl} distributions over strong fronts, weak fronts and outside of fronts.
We assessed three metrics, the local enhancement of al{chl} over fronts, the global enhancement of \al{chl} due to fronts at the scale of the region, and the lag in spring bloom onset due to fronts.
We found that weak fronts lead to a local enhancement of al{chl} weaker than strong fronts, but because they are also more frequent they contribute equally to the regional \al{chl} budget.
We also find the the local enhancement of al{chl} was two to three times larger for the spring bloom than in the oligotrophic subtropical gyre.
We also provide observational evidence that blooms start earlier over fronts, by one to two weeks.
Nevertheless our results suggest that the spectacular impact of fronts at the local scale may be misleading, considering their impact on a regional scale budget remains limited.
\end{abstract}

@@

Despite numerous local observations and a strong theoretical basis for these processes \citep[e.g. recent studies by][]{marrec_2018, little_2018, verneil_2019, ruiz_2019, uchida_2020, kessouri_2020, tzortzis_2021}, their integrated contribution at the scale of regional biomes is still largely unknown.
Indeed, as most fronts form, move and dissipate continuously on time scales of days to weeks, they are particularly difficult to sample, and this limitation is reinforced by the fact that only a limited number of fronts can be observed with in situ field observations.
Thus satellite-derived estimates of al{chl} (hereafter \as{chl}) although limited to the surface and an imperfect proxy for phytoplankton biomass, are the only data that allow to track the impact of fronts synoptically over large areas.
A first attempt to assess the contribution of fine scales to regional satellite \as{chl} budgets was based on a geostatistical analysis derived from data at 9km resolution \citep{doney_2003}, extended later in \citet{glover_2018}, with which they examined the change in spatial variance with distance.
This methodology was too coarse to reveal the impact of submesoscales but confirmed the important role of mesoscale eddies in stirring large scale gradients of phytoplankton abundance.
The role of submesoscales has been assessed with three different methods.
\citet{guo_2019} combined ocean color data with altimetry and drifting floats, and estimated that, over subtropical gyres of the global ocean, the respective contributions of mesoscales and submesoscales to high \as{chl} anomalies were comparable in magnitude.
\citet{keerthi_2022} proposed an approach based on deconvolution of local \as{chl} time series into different timescales; they observed that sub-seasonal time scales contributed roughly 30\% of the total satellite \as{chl} variance and were associated with small (\textless\ 100km) spacial scales.
Finally, the most quantitative approach was proposed by \citet{liu_2016}, which they applied to the North Pacific Subtropical Gyre.
They detected sea-surface temperature (SST) fronts by computing an index that measures the local heterogeneity of the SST field from satellite SST data.
This allowed them to compare satellite \as{chl} values over areas impacted by fronts (characterized by a large value of the heterogeneity index) with values over areas that were not impacted.
They found that the increase in \as{chl} over the fronts was negligible in summer but reached almost 40\% in winter.

Here we built on this last approach, and quantify the excess \as{chl} due to the presence of fronts at the scale of biomes.
This more global quantification will depend on several factors, since the local contribution of fronts to \as{chl} depends on many factors.
First it depends on how efficient fronts are at supplying nutrients, which itself depends on how deep the fronts reach into the nutricline, and of the seasonality of this supply.
It should be noted that the overall efficiency of fronts has been questioned \citep{levy_2018}; firstly because submesoscale fronts are more numerous in winter when convective nutrients input is also greatest; and secondly because the submesoscale vertical currents are often trapped within the mixed-layer, and may not reach the nutricline, which is often well below the base of the boundary mixed-layer.
Second, the overall contribution of fronts will differ between biomes, with submesoscale vertical advection of nutrients likely to be more important in oligotrophic biomes where other nutrient supply routes are scarce, and submesoscale restratification in blooming biomes.
Finally, the contribution of fronts will depend on their spatio-temporal footprint, which also varies seasonally \citep{callies_2015} and regionally \citep{mauzole_2022}.

Thus, more precisely, our intention is to explore and quantify how the contribution of fronts to biome-scale \as{chl} varies in three contrasted biomes, ranging from subtropical to subpolar, varies along the year, and varies with the occurrence and strength of fronts.
We focus our analysis on the North-Atlantic region surrounding the Gulf Stream, where multiple biomes and fronts of different strengths are found in a limited geographical area \citep{bock_2022} with strong seasonality.
In the south, our study area encompasses part of the North Atlantic subtropical gyre, characterized by an oligotrophic regime, year-long low productivity.
In the north, north of the Gulf Stream jet, is a more productive subpolar regime characterized by a recurrent spring bloom.
In between, there is a moderately productive regime, with maximum productivity in winter.
Another feature that makes this study area particularly relevant is that it has different types of fronts.
On the one hand, there are two strong persistent fronts, the Gulf Stream and the shelf-break front, which are held in place by topography and atmospheric circulation, and which are both associated with strong and deep-reaching vertical circulations \citep{flagg_2006, liao_2022}.
But there are also more ephemeral and weaker fronts, created by mesoscale strain that are continuously forming at more random locations \citep{drushka_2019, sanchez-rios_2020}.
We use satellite data of \as{chl} and SST, and extend on the approach of \citet{liu_2016}, to distinguish between persistent and ephemeral fronts.
We evaluate the impact of both types of fronts on \as{chl} on the basis of three indicators, the excess (or deficit) \as{chl} over fronts at the local scale of the front, the surplus \as{chl} attributable to fronts at the scale of regional biomes, and the change in the timing of the \as{chl} spring bloom over fronts.

@@

Our approach combines daily satellite SST data, which are used to detect fronts and sort them by their strength, with daily satellite surface \as{chl}, from which we derive anomalies over fronts.
Our region of interest is the North Atlantic from 15°N to 55°N, and from 40°W to the North American shelf break (Fig.~\ref{fig:zone-separation}).
This region covers three biomes, a more oligotrophic one in the south, a more productive one in the north where a spring bloom occurs, and an intermediate biome between the two, described below.
We will refer to them in the following as permanent subtropical biome (PSB), seasonal subtropical biome (SSB) and subpolar biome (PB), moving from south to north.
They are separated from south to north by a fixed boundary at 32°N and by a variable boundary near 40°N dynamically set at each time step to the instantaneous position of the Gulf Stream.

@@

For \as{chl}, we used the L3 product distributed by ACRI-ST over the period 2000--2020, generated by Copernicus-GlobColour, constructed with data from different sensors (SeaWIFS, MODIS Aqua \& Terra, MERIS, VIIRS-SNPP \& JPSS1, OLCI-S3A \& S3B) merged and reprocessed, available daily at 4km resolution \citep{chl}.

@@

The three biomes in our domain correspond to well known production regimes which have been described previously, although sometimes with different names.
The region is characterized by the presence of a large-scale north-south gradient in \as{chl}.
All pixels where water depth is less than 1500m (red isobath in Fig.~\ref{fig:zone-separation}) are masked to exclude the continental shelf.

The permanently oligotrophic regime to the south of our study area, characterized by warm waters and low \as{chl} (Fig.~\ref{fig:zone-separation}), is known as the subtropical gyre permanently stratified biome \citep{sarmiento_2004} or permanent deep \as{chl} maximum biome \citep{bock_2022}.
Their is no clear physical boundary to the northern limit of this permanent subtropical biome, so we have chosen the latitudinal limit of 32°N to delineate it, which roughly corresponds to the \(0.1 mg m^{-3}\) \as{chl} isocontour in annual mean \as{chl}.

North of 32°N, the seasonal subtropical biome is also mainly oligotrophic, with intermediate levels of \as{chl} and temperature, and characterized by slightly increased productivity in winter.
This second biome is known as the subtropical gyre seasonally stratified biome \citep{sarmiento_2004}.
It is bounded to the north by the meanders of the Gulf Stream jet.
The Gulf Stream jet conveys warm, salty waters poleward along the Florida coast up to Cape Hatteras (35°N), where the jet separates from the continental shelf and meanders essentially zonally.
The north wall of the jet, so called because of its steep temperature gradient, marks the sharp, sinuous and unsteady northern limit of this biome.

@@

The position of the north wall of the Gulf stream, that delimits the subpolar and seasonal subtropical biomes (black meandering contour in Fig.~\ref{fig:zone-separation}a-b), is determined at each daily time step by thresholding the daily SST map.
The daily threshold values (black vertical line in Fig.~\ref{fig:zone-separation}c) are determined from the daily SST distributions above 32°N (blue line, continued by the yellow line in Fig.~\ref{fig:zone-separation}c).
Indeed, the Gulf Stream is easily identifiable in this distribution as it is manifested by a temperature peak (yellow line in Fig.~\ref{fig:zone-separation}c).
To detect the start of this peak, which marks our boundary, we fit the peak with a Gaussian, and define the threshold temperature as the mean minus twice the standard deviation.
Then, the threshold temperature time series is median filtered over an 8-day window to eliminate spurious detection anomalies.
This separation method ensures quasi-unimodal \as{chl} distributions within each biome (Fig.~\ref{fig:zone-separation}d).

@@

\begin{figure*}
  \includegraphics[width=12cm]{zone_separation.pdf}
  \caption{
    Delimitation of the three biomes in the Gulf stream extension region: the Permanent Subtropical Biome (PSB, south of the dashed line at 32°N), the Seasonal Subtropical Biome (SSB, between 32°N and the meandering Gulf stream northern wall on that day marked with the black contour), and the Subpolar Biome (PB, north of the Gulf stream northern wall).
    (a) SST and (b) \as{chl} snapshots on the 22 April 2007 (with data masked by clouds in white), (c) SST and (d) \as{chl}  distribution within each biome for the same day (PB:\@blue, SSB:\@yellow, PSB:\@red).
    The black line in (c) shows the SST threshold value detected to delimit the Gulf Stream northern wall (see methods section).
    The x-axis of the distributions correspond to the x-axis scale of the corresponding color bars.
    The red line follows the 1500m isobath.
    Data on the continental shelf (< 1500m) is not considered here and have been masked.
  }%
  \label{fig:zone-separation}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=12cm]{front_example.pdf}
  \caption{
    The SST, \as{chl}, and heterogeneity index (HI) of a front on the 7 July 2007.
    The plain and dashed contours correspond to HI values of 5 and 10.
    This front is categorized as weak.
    \as{chl} are elevated inside the front.
  }%
  \label{fig:zoom}
\end{figure*}

\subsection{Quantification of the impact of fronts on \as{chl}}

We used three metrics to quantify the impact of fronts on \as{chl}, the excess \as{chl}  E for the local scale (scale of fronts), the surplus \as{chl}  S for the global scale (scale of biomes) and the lag in onset day L for the bloom timing.

In order to quantify how \as{chl} is affected by fronts at the local scale, the metrics E is based on the comparison between the distributions of \as{chl} over fronts and the distributions of \as{chl} outside of fronts.
Distributions were computed over 8-day windows to limit the influence of particularly cloudy days.
They were computed within each biome, and over latitudinal bands of width 5°, to minimize the influence of the large-scale north south gradient in \as{chl}.
Distributions were compared in terms of their median value, but using mean values yielded similar results (not shown).
We defined and computed the local excess \as{chl}, E (expressed in per cent), as the median value over fronts minus the median value in the background, divided by the median value in the background, for each distribution.
We repeated this for weak and strong fronts.

To quantify the large-scale impact of fronts on \as{chl} at the scale of biomes, we computed the biome surplus of \as{chl} which we defined and computed as the relative difference (expressed in per cent) between the mean \as{chl} over the entire biome (MT), and the mean \as{chl} over the background (MB), \(S = (\mathrm{MT}-\mathrm{MB}) / \mathrm{MB}\).
Thus the surplus S measures the extra quantity of \as{chl} at the scale of the biome (\(\mathrm{MT} - \mathrm{MB}\)), relative to what would be the situation in the absence of fronts (MB), and thus accounts for the local increase over fronts (E), but also for the proportion of fronts in a given biome.
To better understand the meaning of the surplus, let us consider the simplified case where \as{chl} is homogeneously doubled over fronts compared to the background value, i.e.\ when E = 100\%; in that case, the surplus \as{chl} is 50\% if there are 50\% of fronts, and is 1\% if there are 1\% of fronts.
Note that the computation of the local excess E is based on median values because it relies on the comparison of distributions, while for the computation of the biome surplus S, we used mean values in order to be conservative.

Finally, the time series of the \as{chl} median in the subpolar biome is characterized by a spring bloom, of which we measured the timing (onset date) both in the fronts and in the background.
Because the spring bloom onsets propagates from south to north, these timings were inferred over latitudinal band of width 5°.
To extract the onset date, we filtered the \as{chl} median time series with a low-pass Butterworth filter of order 2 and cutting frequency \(1/20 \textrm{days}^{-1}\).
The filtered time-series displayed strong variations in their phenology from year to year, but a bloom was always discernible.
We considered data from February to July, which allowed us to isolate the spring bloom and exclude the autumn bloom.
First, we detected the maximum value of \as{chl} in this time window, then defined the bloom onset as the time of maximum \as{chl} derivative prior to the time of maximum \as{chl}.
To estimate the uncertainty in this evaluation, we computed the standard deviation of all days for which the \as{chl} derivative was above 90\% of its maximum value.
Finally, we averaged the yearly values of lag L in onset days over the 20 years range by computing a weighted average and standard deviation of the difference between values in fronts and in background for each latitudinal band, with the weights equal to the inverse of the standard deviations for each year.

@@

\begin{figure*}
  \includegraphics[width=12cm]{chl_vs_hi.pdf}
  \caption{
    Normalized distribution of the Heterogeneity Index (HI, red dashed line) within each biome, and distribution of \as{chl} as a function of HI (representing front strength), over the full period 2000--2020.
    Shown are the median value of the \as{chl} distributions (solid black line), and 1st and 3rd quartiles (dashed lines).
    Note that 0.5\% of pixels have outstanding large HI values and are not included here.
  }%
  \label{fig:chl-vs-hi}
\end{figure*}

@@

\subsection{Local impact of fronts on \as{chl}}

The local impact of fronts on \as{chl} is illustrated by the example shown in Fig.~\ref{fig:zoom}, where the highest values of \as{chl} are found within the HI contour delimiting fronts.
It should be noted, however, that in this example as in other similar examples, the localization of the highest \as{chl} do not perfectly coincide with elevated values of HI;\ they are places where the HI is large and \as{chl} is small, and there are also places with elevated patches of \as{chl} outside of HI contours.

Nevertheless, it appears that on average, \as{chl} is affected by the presence of fronts, and depends on their strength.
In order to measure that effect, we computed for each biome the \as{chl} distribution sorted by bins of HI (of width 0.1) for the whole time series (Fig.~\ref{fig:chl-vs-hi}).
For low values of the heterogeneity index, these distributions are representative of background conditions and reflect the expected differences between biomes: the median \as{chl} is lowest (\(0.05 mg m^{-3}\)) in the permanent subtropical biome, intermediate in the seasonal subtropical biome (\(0.1 mg m^{-3}\)), and highest in the subpolar biome (\(0.2 mg m^{-3}\)).
The \as{chl} variability along the year and within the biome (i.e.\ the width of the distribution, highlighted by grey shading in Fig.~\ref{fig:chl-vs-hi}) is larger moving northwards, because of the higher seasonal variability.

Importantly, in all 3 biomes, \as{chl} values increase with HI, and are also more dispersed as HI increases (Fig.~\ref{fig:chl-vs-hi}).
This suggests that the distinction between background, weak and strong fronts is somehow artificial and that the observed changes in \as{chl} are overall rather continuously dependent on the value of HI.\@
Nevertheless, our partitioning into background, weak and strong fronts remains meaningful because it enables us to isolate the most permanent fronts from the ephemeral ones, and we will therefore retain it in the following.
